---
title: "GAM evaluation - St. Lucie Estuary"
author: "Marcus Beck"
output: 
  html_document:
  toc: yes
self_contained: yes
---

```{r setup, echo = F}
library(knitr)
opts_chunk$set(warning = F, message = F)
```

## Observed data

Observed log-chlorophyll at representative station for the St. Lucie Estuary  
```{r proc, fig.width = 7, fig.height = 4}
library(tidyverse)
library(lubridate)
library(mgcv)  
library(plotly)
library(WRTDStidal)
library(gridExtra)
source('R/funcs.R')

# flow data, left moving window average of 30 days
data(sl_fldat)
fl_dat <- sl_fldat %>% 
  rename(date = Date) %>% 
  mutate(
    qsm = stats::filter(q, rep(1, 30)/30, sides = 1, method = 'convolution')
  )

# format the data to model
data(sl_dat)
sl_mod <- sl_dat %>%
  rename(date = Date) %>% 
  mutate(
    doy = yday(date), 
    dec_time = decimal_date(date), 
    yr = year(date),
    yr = factor(yr),
    mo = month(date, label = T)
  ) %>% 
  left_join(fl_dat, by = 'date') %>% 
  mutate(
    flo = log(qsm),
    chl = log(1 + chl)
    ) %>% 
  select(-q, -qsm, -sal)

# plot, all
p <- ggplot(sl_mod, aes(x = date, y = chl)) + 
  geom_line() +
  theme_bw() 
ggplotly(p)

# boxplot, by years
p <- ggplot(sl_mod, aes(x = yr, y = chl)) + 
  geom_boxplot() + 
  theme_bw()
ggplotly(p)

# boxplot, by month
p <- ggplot(sl_mod, aes(x = mo, y = chl)) + 
  geom_boxplot() + 
  theme_bw()
ggplotly(p)
```

## GAMs with annual, seasonal trends

Some simple GAMs to explore annual, seasonal trends.
```{r mods}
# smooths to evaluate
smths <- c(
  "s(dec_time, bs = 'tp')",  
  "s(doy, bs = 'cc')",
  "te(dec_time, doy, bs = c('tp', 'cc'))"
)

# get all combinations of smoothers to model, one to many
frms <- list()
for(i in seq_along(smths)){
  
 frm <- combn(smths, i) %>%
    apply(2, function(x){
      paste(x, collapse = ' + ') %>% 
        paste('chl ~ ', .) %>% 
        formula
    }) 
  
 frms <- c(frms, frm)
 
}

# create models from smooth formula combinations
mods <- map(frms, function(frm){
  
  gam(frm, 
    knots = list(doy = c(1, 366)),
    data = sl_mod, 
    na.action = na.exclude
  )

})
names(mods) <- paste0('mod', seq_along(mods))
```

Summary stats of annual, seasonal models:
```{r modsum, results = 'asis'}
# smoother stats of GAMs
map(mods, ~ summary(.x)$s.table %>% data.frame %>% rownames_to_column('smoother')) %>% 
  enframe %>% 
  unnest %>% 
  kable(digits = 2)

# summary stats of GAMs
map(mods, ~ data.frame(
    AIC = AIC(.x), 
    R2 = summary(.x)$r.sq)) %>% 
  enframe %>% 
  unnest %>% 
  kable(digits  = 2)
```

```{r predplo, fig.width = 7, fig.height = 4, eval = T}
# prediction data
pred_dat <- crossing(
    yr = seq(floor(min(sl_mod$dec_time)), ceiling(max(sl_mod$dec_time))),
    doy = seq(1, 365, by = 5)
  ) %>%
  unite('date', yr, doy, sep = '-', remove = F) %>% 
  mutate(
    date = as.Date(date, format = '%Y-%j'),
    dec_time = decimal_date(date),
    mo = month(date, label = TRUE), 
    yr = year(date)
  ) %>% 
  left_join(., fl_dat[, c('date', 'qsm')]) %>% 
  mutate(flo = log(qsm)) %>% 
  select(-qsm)

# predictions
sl_res <- map(mods, function(x){
  pred_dat %>% 
    mutate(
      pred = predict(x, newdata = pred_dat)
    )
  }) %>% 
  enframe('mods') %>% 
  unnest

# plot
p <- ggplot(sl_res, aes(x = date)) + 
  geom_point(data = sl_mod, aes(y = chl), size = 0.5) + 
  geom_line(aes(y = pred, colour = mods)) + 
  theme_bw() + 
  theme(
    legend.position = 'top', 
    legend.title = element_blank()
    )
ggplotly(p)
```

```{r predplo2, message = F, warning = F, fig.width = 7, fig.height = 9, eval = T}
# plot
p <- ggplot(sl_res, aes(x = doy, group = factor(yr), colour = yr)) + 
  geom_line(aes(y = pred)) + 
  theme_bw() + 
  theme(
    legend.position = 'top', 
    legend.title = element_blank()
    ) + 
  facet_wrap(~ mods, ncol = 2)
ggplotly(p)
```

## GAMs with annual, seasonal, and flow trends

Adding flow data to the model:
```{r flomod, eval = F}
# smooths to evaluate
smths <- c(
  "s(dec_time, bs = 'tp')",  
  "s(doy, bs = 'cc')",
  "s(flo, bs = 'tp')",
  "te(flo, doy, bs = c('tp', 'cc'))", 
  "te(flo, dec_time, bs = c('tp', 'tp'))",
  "te(dec_time, doy, bs = c('tp', 'cc'))",
  "te(dec_time, doy, flo, bs = c('tp', 'cc', 'tp'))"
)

# get all combinations of smoothers to model, one to many
frms <- list()
for(i in seq_along(smths)){
  
 frm <- combn(smths, i) %>%
    apply(2, function(x){
      paste(x, collapse = ' + ') %>% 
        paste('chl ~ ', .) %>% 
        formula
    }) 
  
 frms <- c(frms, frm)
 
}

# create models from smooth formula combinations
mods2 <- map(frms, function(frm){
  
  gam(frm, 
    knots = list(doy = c(1, 366)),
    data = sl_mod, 
    na.action = na.exclude
  )

})
names(mods2) <- paste0('mod', seq_along(mods2))
save(mods2, file = 'data/sl_mods2.RData', compress = 'xz')
```

Summary stats of best year/season model, year/season/flow model:
```{r modsum2, results = 'asis', eval = T}
data(sl_mods2)
# best model with only season, year
best1 <- map(mods, ~ summary(.x)$r.sq) %>% 
  unlist %>% 
  which.max %>% 
  mods[[.]]

# best model with season, year, flow
best2 <- map(mods2, ~ summary(.x)$r.sq) %>% 
  unlist %>% 
  which.max %>% 
  mods2[[.]] 

best <- list(best1 = best1, best2 = best2)

# smoother stats of GAMs
map(best, ~ summary(.x)$s.table %>% data.frame %>% rownames_to_column('smoother')) %>% 
  enframe %>% 
  unnest %>% 
  kable(digits  = 2)

# summary stats of GAMs
map(best, ~ data.frame(
    AIC = AIC(.x), 
    R2 = summary(.x)$r.sq)) %>% 
  enframe %>% 
  unnest %>% 
  kable(digits  = 2)
```

```{r predplo3, fig.width = 7, fig.height = 4, eval = T}
# predictions
sl_res2 <- map(best, function(x){
  pred_dat %>% 
    mutate(
      pred = predict(x, newdata = pred_dat)
    )
  }) %>% 
  enframe('mods') %>% 
  unnest

# plot
p <- ggplot(sl_res2, aes(x = date)) + 
  geom_point(data = sl_mod, aes(y = chl), size = 0.5) + 
  geom_line(aes(y = pred, colour = mods)) + 
  theme_bw() + 
  theme(
    legend.position = 'top', 
    legend.title = element_blank()
    )
ggplotly(p)
```

```{r predplo4, fig.width = 4, fig.height = 14, eval = T}
ptheme <- theme(
  axis.title.x = element_blank(), 
  axis.title.y = element_blank()
)
cols <- 'Spectral'
pb1 <- dynagam(best1, pred_dat, ncol = 1, col_vec = cols) + 
  ptheme + 
  theme(legend.position = 'none') +
  ggtitle('Best 1')
pb2 <- dynagam(best2, pred_dat, ncol = 1, col_vec = cols) + 
  ptheme + 
  ggtitle('Best2')
pleg <- g_legend(pb2)
pb2 <- pb2 + 
  theme(legend.position = 'none')

grid.arrange(
  pleg, 
  arrangeGrob(pb1, pb2, ncol = 2, bottom = 'lnQ', left = 'lnchl'), 
  ncol = 1, 
  heights = c(0.1, 1)
)
```

## Adding nitrogen and turbidity covariates

```{r mods3, eval = T}
# formula for best annual, seasonal, flow model
strt <- best2$formula %>% 
  as.character

smths <- c(
  "s(nh, bs = 'tp')",  
  "s(tss, bs = 'tp')",
  "te(nh, tss, bs = c('tp', 'tp'))"
)

# get all combinations of smoothers to model, one to many
frmstab <- list()
frms <- list()
for(i in seq_along(smths)){
  
  # for the summary table  
  frmtab <- combn(smths, i) %>%
    apply(2, function(x){
      paste(x, collapse = ' + ')
      })
  
  # for the model
  frm <- sapply(frmtab, function(x){  
        paste('chl ~', strt[3], '+', x) %>% 
          formula
        })
  
  frmstab <- c(frmstab, frmtab)
  frms <- c(frms, frm)
 
}

# create models from smooth formula combinations
mods3 <- map(frms, function(frm){
  
  gam(frm, 
    knots = list(doy = c(1, 366)),
    data = sl_mod, 
    na.action = na.exclude
  )

})
names(mods3) <- paste0('mod', seq_along(mods3))
```

Summary of all nutrient, turbidity models
```{r modsum3, results = 'asis', eval = T}
# smoother stats of GAMs
map(mods3, ~ summary(.x)$s.table %>% data.frame %>% rownames_to_column('smoother')) %>% 
  enframe %>% 
  unnest %>% 
  kable(digits  = 2)

# summary stats of GAMs
map(mods3, ~ data.frame(
    AIC = AIC(.x), 
    R2 = summary(.x)$r.sq)) %>% 
  enframe %>% 
  unnest %>% 
  mutate(smth_added = frmstab) %>% 
  select(name, smth_added, everything()) %>% 
  kable(digits  = 2)
```

Summary stats of best three three models:
```{r modsum4, results = 'asis'}
# best model with season, year, flow
best3 <- map(mods3, ~ summary(.x)$r.sq) %>% 
  unlist %>% 
  which.max %>% 
  mods3[[.]] 

best <- list(best1 = best1, best2 = best2, best3 = best3)

# smoother stats of GAMs
map(best, ~ summary(.x)$s.table %>% data.frame %>% rownames_to_column('smoother')) %>% 
  enframe %>% 
  unnest %>% 
  kable(digits  = 2)

# summary stats of GAMs
map(best, ~ data.frame(
    AIC = AIC(.x), 
    R2 = summary(.x)$r.sq)) %>% 
  enframe %>% 
  unnest %>% 
  kable(digits  = 2)
```

```{r predplo5, fig.width = 7, fig.height = 4, eval = T}
# predictions
sl_res3 <- map(best, function(x){
  sl_mod %>% 
    mutate(
      pred = predict(x)
    )
  }) %>% 
  enframe('mods') %>% 
  unnest

# plot
p <- ggplot(sl_res3, aes(x = date)) + 
  geom_point(data = sl_mod, aes(y = chl), size = 0.5) + 
  geom_line(aes(y = pred, colour = mods)) + 
  theme_bw() + 
  theme(
    legend.position = 'top', 
    legend.title = element_blank()
    )
ggplotly(p)
```